Predicting the effects of character displacement on food-web dynamics

### 2019-10-30 manuscript_ECD_model.tex
final edits to main text for bioRxiv submission

\documentclass[11pt,]{article}
\usepackage{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Matthew A. Barbour\^{}\{1,2,\textbackslash{}ast\}},
            pdftitle={Predicting the effects of character displacement on food-web dynamics},
            colorlinks=true,
            citecolor=blue,
            urlcolor=black,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

  \title{Predicting the effects of character displacement on food-web dynamics}
    \author{Matthew A. Barbour\(^{1,2,\ast}\)}
    \date{}

%%% From AmNat_MS_template.tex
%\documentclass[11pt]{article}
%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
%\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
%\usepackage[utf8]{inputenc}
\usepackage{lineno}

@@

\maketitle

\noindent 1. University of British Columbia, Department of Zoology,
Vancouver, BC V6T 1Z4, Canada;

\noindent 2. University of Zurich, Department of Evolutionary Biology
and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich,
Switzerland;

@@

\emph{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4. All
figures should be printed in color.

@@

\emph{Keywords}: competition; eco-evolutionary dynamics;
consumer-resource interactions; adaptive radiation; ecological community

\emph{Manuscript type}: Note.

\footnotesize Prepared using an \emph{Am. Nat.} inspired \LaTeX{}
template for Rmarkdown. \normalsize

@@

@@

Ecological character displacement is thought to be a key evolutionary
process generating biodiversity (Dolph Schluter 2000; D. W. Pfennig and
Pfennig 2010; but see Stuart and Losos 2013). Ecological character
displacement describes the ``process of phenotypic evolution in a
species generated or maintained by {[}exploitative{]} resource
competition with one or more coexisting species'' (Dolph Schluter 2000).
Over the past 40 years, a large body of theoretical (e.g. Lawlor and
Smith 1976; P A Abrams 1986; Doebeli 1996; Taper and Chase 1985) and
empirical (reviewed in: Dolph Schluter 2000; Dayan and Simberloff 2005;
Stuart and Losos 2013) work has been generated to understand the
scenarios under which exploitative competition for resources leads to
the divergence of consumer foraging traits. The general result emerging
from these studies is that, if resources are nutritionally substitutable
(Peter A Abrams 1987; Fox and Vasseur 2008) and there is no other strong
source of density dependence acting on consumers (P A Abrams 1986), then
resource competition drives the divergence of consumer foraging traits
{[}Lawlor and Smith (1976); Taper and Chase (1985)). This process is not
simply driven by ecological differences, but creates an eco-evolutionary
feedback that drives further differentiation. This key insight was made
by theoretical models that explicitly included resource dynamics as a
mediator of competition in driving evolutionary change (Lawlor and Smith
1976; P A Abrams 1986; Taper and Chase 1985).

Although explicit models of resources gave key insights to the evolution
of character displacement, the ecological feedback onto
consumer-resource dynamics have received surprisingly little attention.
This is likely because character displacement has been primarily studied
through the lens of competition and coexistence theory {[}Lawlor1976;
Germain2018{]}. For example, early theoretical work by Lawlor and Smith
(1976) showed that ecological character displacement promotes
coexistence by favoring specialized consumers that experience reduced
interspecific competition. On the other hand, specialization often
increases a consumer's ability to capture a specific resource. A well
known consequence of increased foraging efficiency is resource
suppression, which if sufficient enough, can generate oscillations, and
hence less stable, consumer-resource dynamics {[}Rosenzweig 1971;
Murdoch et al. 2003; McCann 2011{]}. NEED A LAST SENTENCE TO WRAP IT UP,
HIGHLIGHTING HOW A FOOD-WEB PERSPECTIVE MAY YIELD NEW ECOLOGICAL
INSIGHTS AND GIVE A CONTRASTING PERSPECTIVE TO COEXISTENCE THEORY.
PERHAPS A I NEED TO INCLUDE A SENTENCE MID-WAY THROUGH SUGGESTING THAT A
FOOD-WEB PERSPECTIVE MAY GIVE QUALITATIVELY DIFFERENT INSIGHTS. IT MAY
BE NICE TO BRING IN THE SPATIAL COUPLING WORK OF MCCANN SUGGESTING THAT
SPECIALIZATION CAN COUPLE FOOD-WEBS, BUT THIS WAS ONLY WITH ONE TOP
PREDATOR (I.E. NO CONSUMERS\ldots{}LOOK AT OLD TEXT FOR WORDING
SUGGESTIONS.)

Here, I address this knowledge gap by building a mathematical model to
examine how ecological character displacement affects consumer-resource
dynamics in a food-web context. I address two questions: (1) How does
ecological character displacement affect resource abundances? (2) How
does character displacement affect food-web stability? To test the
generality of these ecological effects, I examined three different
foraging scenarios: (i) resources occur in the same habitat and
consumers exhibit a linear functional response (e.g. MacArthur (1972);
Lawlor and Smith (1976)); (ii) resources occur in different habitats and
consumers exhibit a linear functional response (e.g. Lawlor and Smith
(1976)); (iii) resources occur in distinct habitats and consumers
exhibit a more realistic functional response (K S McCann, Rasmussen, and
Umbanhowar 2005). I also explore the contingency of these effects to
different assumptions about evolutionary trade-offs in consumer foraging
traits. I focus on the ecological side of these dynamics because the
evolutionary aspects have been well characterized (Lawlor and Smith
1976; Taper and Chase 1985; P A Abrams 1986). I show that the
consequences of character displacement on food-web dynamics are not
always intuitive and depend on the foraging scenario and tradeoff in
consumer traits.

\section{Methods and Results}\label{methods-and-results}

\subsection{Underlying Food-web
Dynamics}\label{underlying-food-web-dynamics}

To examine how ecological character displacement affects
consumer-resource interactions and food-web dynamics, I analyzed a
continuous-time model of two consumers (\(C_1,C_2\)) competing for two
shared resources (\(R_1,R_2\)):

\[\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2,\]
\[\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2,\]
\[\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1,\]
\[\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2,\]

where \(r_i\) represents the intrinsic growth rate of resource \(i\),
\(K_i\) represents the carrying capacity of resource \(i\), \(e_{ji}\)
represents the conversion efficiency of resource \(i\) into consumer
\(j\), and \(m_j\) represents the mortality rate of consumer \(j\).
\(F_{ji}(R_i)\) represents consumer \(j\)'s feeding rate
(i.e.~functional response) on resource \(i\). This model is a useful
characterization of a scenario where consumers compete for two distinct
resources (e.g.~zooplankton and benthic invertebrates) rather than if
resources are better characterized by a continuous trait distribution
(e.g.~seed size, see Taper and Chase (1985) for an example). In the
scenario where consumer feeding rates increase linearly with resource
density, such that \(F_{ji}(R_i)=a_{ji}R_i\) where \(a_{ji}\) is the
attack rate of consumer \(j\) on resource \(i\), then this model becomes
MacArthur's model of resource competition (MacArthur (1972)). The
evolution of consumer attack rates in this model (and several
extensions) have been analyzed in detail (Lawlor and Smith (1976); P A
Abrams (1986)), with the general result being divergent character
displacement. I say consumers have undergone divergent character
displacement if there evolved attack rates are more specialized, defined
as \(|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|\), when evolving with a
competing consumer. I focus here on the ecological consequences of this
adaptive divergence.

To investigate the ecological consequences, I studied the effect of
adaptive divergence in consumers on resource densities and the stability
of the entire food web at equilibrium. An equilibrium is reached when
the four rates of change in the above equations are 0, and solving the
system at this point gives equilibrium densities for each resource
(\(\hat R_1,\hat R_2\)) and consumer (\(\hat C_1,\hat C_2\)). At the
coexistence equilibrium (i.e., where
\(\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0\)), I calculated resource
densities and performed a linear stability analysis (Otto and Day 2007).
This stability analysis derives the dominant (largest in absolute value)
eigenvalue, \(\lambda\), which determines whether (and how readily) the
food web, when perturbed a small amount, will return to equilibrium (see
supplementary material for derivation). Together, these two measures
indicate how ecological character displacement affects resource
densities and food-web stability.

I examine the effects of this inevitable divergence in attack rates on
food-web dynamics by (i) deriving analytical expressions for the
relationship between attack rates and food-web dynamics and (ii)
simulating the effects of competition on the eco-evolutionary dynamics
of consumers and resources. To gain analytical insight, I assume that
resources are equivalent (\(r=r_1=r_2\) and \(K=K_1=K_2\)) as well as
consumers (\(e=e_{11}=e_{12}=e_{21}=e_{22}\); \(m=m_1=m_2\)), except
that consumer attack rates are mirror images of each other
(\(a_{ii}=a_{11}=a_{22}\); \(a_{ij}=a_{12}=a_{21}\)). While this
scenario is arguably simplistic, it allows me to isolate the effects of
ecological character displacement.

To simulate eco-evolutionary dynamics, I used an Adaptive Dynamics
approach. At each evolutionary time step, I created a mutant consumer by
randomly choosing a consumer and modifying its attack rate on one
resource by either subtracting or adding a small constant (0.001 in the
following simulations) with equal probability. The mutant's attack rate
on the other resource was determined by a tradeoff, such that
\(\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1\), where \(A\) is the
total investment in attack rates and \(n\) describes the shape of the
tradeoff (Sargent and Otto 2006). This function has the useful property
that it differentiates between cases where intermediate combinations of
\(a_{ii}\) and \(a_{ij}\) are higher, on average, than the extremes
(when \(n>1\), green line in Fig. \ref{fig:Tradeoff}) or, conversely,
where the two extremes are higher, on average, than intermediate
investments (when \(n<1\), orange line Fig. \ref{fig:Tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{ii}\) and \(a_{ij}\) have the same total attack rate (blue line in
Fig. \ref{fig:Tradeoff}). Assuming the mutant consumer was rare, I then
determined whether the mutant had higher relative fitness than the
resident consumer, and thus could invade and replace the resident
consumer. If the mutant was able to invade, I updated the attack rate of
the resident consumer to the mutant attack rate and allowed consumer and
resource dynamics to reach a steady state. I then repeated the
simulation up to 10,000 times, which was sufficient for consumers to
either reach an evolutionary stable strategy (ESS, Smith and Price 1973)
or an evolutionary limit (e.g. \(\frac{a_{ii}}{a_{ii}+a_{ij}}\) is
constrained to a maximum of 1 and minimum of 0).

\begin{figure}
\centering
\includegraphics{manuscript_ECD_model_files/figure-latex/Tradeoff-1.pdf}
\caption{\label{fig:Tradeoff}Tradeoff forms for consumer attack rates.}
\end{figure}

\subsection{Consumers Forage For Both Resources
Simultaneously}\label{consumers-forage-for-both-resources-simultaneously}

When both consumers and resources are present, the density of resources
at equilibrium are equivalent (\(\hat R = \hat R_1 = \hat R_2\)) and are
determined by the following equation (derivation in supplementary
material):

\[\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}\] A key determinant of
resource density in this model is the consumer's total attack rate,
\(a_{ii}+a_{ij}\). Therefore, the ecological consequences of character
displacement depend on how the tradeoff influences the evolution of
consumer attack rates.

My simulations show that the shape of the attack-rate tradeoff
qualitatively affects the relationship between character displacement
and resource density (Fig. \ref{fig:MacArthur_Resources}). For example,
if consumer's are constrained by a linear tradeoff (blue lines), then
there is no net change in total attack rate (Fig.
\ref{fig:MacArthur_Resources}A) and character displacement has no effect
on resource densities (Fig. \ref{fig:MacArthur_Resources}B). If the
tradeoff is concave down (green lines), then resource abundances can
actually increase under character displacement (Fig.
\ref{fig:MacArthur_Resources}B). This is because the total attack rate
of consumers is maximized at intermediate values (\(a_{ii}=a_{ij}\)) and
decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A).
When the tradeoff is concave up (orange lines), character displacement
suppresses resource densities due to the increase in total attack rates
(Fig. \ref{fig:MacArthur_Resources}A,B). Although the equation I derived
for resource density was for the scenario where both consumers and both
resources are present, it accurately predicts the density of resources
when a single consumer reaches its ESS (triangles on respective colored
lines in Fig. \ref{fig:MacArthur_Resources}B). This is because a single
consumer evolves to be a generalist that has equal attack rates on each
resource (triangles at 0.5 along x-axis in Fig.
\ref{fig:MacArthur_Resources}A), resulting in equivalent resource
densities.

The effect of character displacement on resource density closely matches
its impact on food-web stability (Fig. \ref{fig:Stability}). For
example, when character displacement results in resource suppression
(orange), there is a corresponding decrease in stability. Similarly, if
character displacement does not influence resource densities (blue),
then there is no corresponding effect on stability. Curiously, this
match between impact on resources and stability breakdowns with the
concave down tradeoff (green).

\begin{figure}
\centering
\includegraphics{manuscript_ECD_model_files/figure-latex/MacArthur_Resources-1.pdf}
\caption{\label{fig:MacArthur_Resources}Effect of character displacement
on total attack rates (A) and resource densities (B) when consumers can
forage for both resources simultaneously.}
\end{figure}

\includegraphics{Figure_ECD_MacArthur.pdf}

\subsection{\texorpdfstring{Consumers \emph{cannot} forage for both
resources
simulanteously}{Consumers cannot forage for both resources simulanteously}}\label{consumers-cannot-forage-for-both-resources-simulanteously}

The only character displacement model that I am aware of that modeled
resources in different habitats was one examined by Lawlor and Smith
(1976). This model takes the same form as the previous model, except now
the consumer's feeding rate takes the form:

\[F_{ji}(R_i)=w_{ji}a_{ji}R_i\]

where \(w_{ji}\) represents the proportion of time consumer \(j\) spends
foraging in a habitat where only resource \(j\) is found (i.e.~habitat
preference). Note that since \(w_{ji}\) is a proportion that
\(w_{jj}=1-w_{ji}\). As with attack rates, I assume that consumer
habitat preferences are mirror images of each other
(\(w=w_{11}=w_{22}\)). In Lawlor and Smith (1976)'s detailed
evolutionary analysis of this model, they again always observe divergent
character displacement.

>>>>

\documentclass[11pt,]{article}
\usepackage[]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=black,
            urlcolor=black,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

%%% From AmNat_MS_template.tex
%\documentclass[11pt]{article}
%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
%\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.5} % 1.7
%\usepackage[utf8]{inputenc}
\usepackage{lineno}

% controlling text size in figure captions
\usepackage{caption}
\captionsetup[figure]{font=small}

@@

\vspace*{0.1cm}

\begin{center} \LARGE Ecological character displacement destabilizes food webs \end{center}

\begin{center} \large Matthew A. Barbour$^{1,2,\ast}$ \normalsize \end{center}

\noindent 1. University of Zurich, Department of Evolutionary Biology
and Environmental Studies, 8057 Zurich, Switzerland;

\noindent 2. University of British Columbia, Department of Zoology,
Vancouver, BC V6T 1Z4, Canada;

@@

\emph{Running title}: Character divergence destabilizes food webs

@@

\emph{Keywords}: competition; eco-evolutionary dynamics;
consumer-resource interactions; adaptation; community stability

@@

Ecological character displacement is an adaptive process that generally
increases phenotypic diversity. Despite the fact that this
diversification is due to an eco-evolutionary feedback between consumers
competing for shared resources, its consequences for food-web dynamics
have not been examined. Here, I study a model of two consumers competing
for two shared resources to examine how character displacement in
consumer attack rates affects resource abundances and the resilience of
food webs to perturbations. I found that character displacement always
strengthened consumer-resource interactions whenever consumers competed
for resources that occurred in different habitats. This increase in
interaction strength resulted in lower resource abundances and less
resilient food webs. This occurred under different evolutionary
tradeoffs and in both simple and more realistic foraging scenarios.
Taken together, my results show that the adaptive process of character
displacement may come with the ecological cost of decreasing food-web
resilience.

@@

Ecological character displacement is an important adaptive process in
generating biodiversity \citep{Schluter2000, Pfennig2010}. This process
is due to ``phenotypic evolution in a species generated or maintained by
{[}exploitative{]} resource competition with one or more coexisting
species'' \citep{Schluter2000}. A large body of theoretical
\citep[e.g.][]{Lawlor1976, Abrams1986, Doebeli1996, Taper1985, McPeek2019}
and empirical \citep[reviewed
in:][]{Schluter2000, Dayan2005, Stuart2013} work has examined which
scenarios lead to phenotypic divergence or convergence of competing
consumers. The general conclusion has been that, if resources are
nutritionally substitutable \citep{Abrams1987, Fox2008} and there is no
other strong source of density dependence acting on consumers
\citep{Abrams1986}, then resource competition drives the adaptive
divergence of competitors \citep{Lawlor1976, Taper1985}. This adaptive
process is not simply a response to static differences in resource
distributions, but creates an eco-evolutionary feedback that drives
further differentiation. This crucial insight was made by theoretical
models that explicitly included resource dynamics as a mediator of
competition in driving evolutionary change
\citep{Lawlor1976, Abrams1986, Taper1985}.

Although models that included resources led to insights to the evolution
of character displacement, the ecological feedback onto
consumer-resource dynamics has received surprisingly little attention.
This is likely because the ecological feedback has been primarily
studied through the lens of coexistence theory
\citep{Lawlor1976, Germain2018, Bassar2017, McPeek2019}. For example,
early theoretical work showed that ecological character displacement
promotes coexistence by favoring specialized consumers that experience
reduced interspecific competition \citep{Lawlor1976}. Yet, this
reduction in interspecific competition may, at the same time, increase
interspecific interactions between specialized consumers and their
resources. Both food-web theory and empirical studies have shown that
increasing the strength of consumer-resource interactions often
suppresses the abundance of resources, which if sufficient enough, can
generate oscillations and less stable consumer-resource dynamics
\citep{Rosenzweig1971, Luckinbill1973, Murdoch2002, Murdoch2003, McCann2011}.
Thus, a food-web perspective, which accounts for both the direct and
indirect effects of consumer-resource interactions, may yield new
insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by studying a mathematical model that
examines how ecological character displacement affects consumer-resource
dynamics in a food-web context. I address two questions: (1) How does
ecological character displacement affect resource abundances? (2) How
does character displacement affect food-web stability? To test the
generality of these effects, I explored different ecological foraging
scenarios and evolutionary tradeoffs in consumer attack rates. I found
that the adaptive process of character displacement often comes with an
ecological cost; resulting in food webs with lower resource availability
and that are less resilient to perturbations.

\section{Material and methods}\label{material-and-methods}

\subsection{Underlying Consumer-Resource
Dynamics}\label{underlying-consumer-resource-dynamics}

To examine how ecological character displacement affects resource
abundances and food-web stability, I analyzed a continuous-time model of
two consumers (\(C_{j=1,2}\)) competing for two shared resources
(\(R_{i=1,2}\)):

\begin{equation} \label{eq:1}
  \begin{split}
     & \frac{dR_1}{dt}=r_1R_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{12}(R_1)C_2 \\
     & \frac{dR_2}{dt}=r_2R_2(1-\frac{R_2}{K_2})-F_{21}(R_2)C_1-F_{22}(R_2)C_2 \\
     & \frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{21}F_{21}(R_2)C_1-m_1C_1 \\
     & \frac{dC_2}{dt}=e_{12}F_{12}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2 \\
  \end{split}
\end{equation}

where \(r_i\) represents the intrinsic growth rate of resource \(i\),
\(K_i\) represents the carrying capacity of resource \(i\), \(e_{ij}\)
represents the conversion efficiency of resource \(i\) into consumer
\(j\), and \(m_j\) represents the mortality rate of consumer \(j\).
\(F_{ij}(R_i)\) represents consumer \(j\)'s feeding rate on resource
\(i\) (i.e.~functional response). This model is a useful
characterization of a scenario where consumers compete for two distinct
resources (e.g.~zooplankton and benthic invertebrates in lakes) rather
than a scenario where resources are better characterized by a continuous
trait distribution (e.g.~seed size, see \citet{Taper1985} for an
example). Importantly, inferences about ecological character
displacement can only be made by comparing food webs with and without a
competing consumer \citep{Schluter1992}. Therefore, I arbitrarily set
\(C_2=0\) to create a food-web without a competing consumer for these
comparisons.

\subsection{Foraging Scenarios}\label{foraging-scenarios}

I studied three different foraging scenarios. In the first, I assume
that consumers can forage for both resources simultaneously (fig.
\ref{fig:foraging_scenarios} top) and their feeding rates increase
linearly with resource abundance, such that:

\begin{equation} \label{eq:2}
  F_{ij}(R_i)=a_{ij}R_i
\end{equation}

where \(a_{ij}\) is the attack rate of consumer \(j\) on resource \(i\).
This first scenario is the starting point for many models of resource
competition \citep{MacArthur1972}; however, it does not reflect many
food webs where consumers are mobile and their foraging behavior links
resources that occur in different habitats \citep{McCann2005}. The
second scenario accounts for this spatial context (fig.
\ref{fig:foraging_scenarios} bottom) and takes the form:

\begin{equation} \label{eq:3}
  F_{ij}(R_i)=w_{ij}a_{ij}R_i
\end{equation}

where \(w_{ij}\) represents the proportion of time consumer \(j\) spends
foraging in a habitat where only resource \(i\) is found (i.e.~habitat
preference). Note that since \(w_{ij}\) is a proportion that
\(w_{1,j}=1-w_{2,j}\). Finally, it is well known that consumer feeding
rates often saturate at high resource abundances
\citep{Holling1959, Rosenzweig1963, Murdoch2003, McCann2011} and that
consumers do not usually spend a fixed proportion of time in a
particular habitat \citep{McCann2005}. The third scenario accounts for
these biological realities and takes the form \citep[derived
by][]{McCann2005}:\\

\begin{equation} \label{eq:4}
  F_{ij}(R_i)=\frac{a_{ij}W_{ij}R_i}{1+a_{1,j}h_{1,j}W_{1,j}R_1+a_{2,j}h_{2,j}W_{2,j}R_2}
\end{equation}

where consumer \(j\)'s feeding rate on resource \(i\) is influenced by
the abundance of each resource; saturates as resource abundances
increase (due to handling times \(h_{ij}\)); and consumer habitat
preferences are modified by the relative abundance of resources, such
that: \(W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}\).

Previous studies have analyzed the evolution of consumer attack rates in
the first two foraging scenarios using an Adaptive Dynamics approach,
with the general result being character divergence
\citep{Lawlor1976, Abrams1986}. I also used an Adaptive Dynamics
approach to analyze character displacement in the third foraging
scenario, and I too observed character divergence (detailed analysis in
Appendix S1 of Supplementary Information). I say consumers have
undergone character divergence if their evolved attack rates are more
specialized when evolving with vs.~without a competing consumer.
Specialization of consumer \(j\) on resource \(1\) is measured as
\(\frac{a_{1,j}}{a_{1,j}+a_{2,j}}\), where a value of 0.5 is a complete
generalist (\(a_{1,j}=a_{2,j}\)), and a value of 1 is a complete
specialist (\(a_{2,j}=0\)). Values less than 0.5 indicate specialization
on the other resource.

\begin{figure}
\centering
\includegraphics{ECD_model_foraging_scenarios}
\caption{\label{fig:foraging_scenarios}\textbf{Ecological foraging
scenarios.} I examined whether the effect of ecological character
displacement on food-web dynamics depended on whether consumers competed
for resources that occurred in the same vs.~different habitats. The
width of each arrow corresponds to initial attack rates
(\emph{a}\textsubscript{ij}) of consumer \emph{j} on resource \emph{i}.
Note that \emph{C}\textsubscript{1} was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{21}), while \emph{C}\textsubscript{2} was a
mirror image, being pre-adapted to \emph{R\textsubscript{2}}
(\emph{a}\textsubscript{22} \textgreater{} \emph{a}\textsubscript{12}).
In each scenario, I assumed consumer feeding rates increased linearly
with resource abundance. I also relax this assumption and consider a
more realistic functional response when resources occurred in different
habitats.}
\end{figure}

\subsection{Food-web Dynamics}\label{food-web-dynamics}

Given that character divergence occurred across these foraging
scenarios, I focus here on its consequences for food-web dynamics. To do
this, I analyzed differences in resource abundances and food-web
stability at equilibrium. An equilibrium is reached when the rates of
change in equation \ref{eq:1} are 0, and solving the system at this
point gives equilibrium abundances for each resource (\(\hat R_i\)) and
each consumer (\(\hat C_j\)). I also compared the local stability of
these food webs using standard methods \citep{Otto2007}. This stability
analysis derives the dominant eigenvalue, \(\lambda_{\text{max}}\), of
the matrix of partial derivatives of all differential equations with
respect to all variables evaluated at equilibrium. If
\(-\lambda_{max}>0\), then the food web will return to equilibrium after
a small perturbation (i.e.~locally stable), with more positive values
indicating a faster return time. If \(-\lambda_{max}<0\), then the food
web is not locally stable.

When possible, I derived analytical expressions for the relationship
between consumer attack rates and food-web dynamics. To do this, I
simplified the model by assuming that resources are equivalent
(\(r=r_i\) and \(K=K_i\)) as well as consumers (\(e=e_{ij}\);
\(h=h_{ij}\); \(m=m_j\)), except that consumer attack rates and their
habitat preferences (if present) are mirror images of each other
(\(a_{11}=a_{22}\); \(a_{12}=a_{21}\); \(w_{11}=w_{22}\)). Note that I
arbitrarily set \(C_1\) as being pre-adapted to \(R_1\)
(\(a_{11} > a_{21}\); \(w_{11} > 0.5\)), and therefore \(C_2\) was
pre-adapted to \(R_2\). Controlling for other sources of variability
allowed me to isolate the general effects of character divergence. All
mathematical derivations were conducted in Mathematica
\citep{Mathematica} and are provided in the Supplementary Information
(Appendix S1-3).

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0). Unless otherwise noted, I conducted
simulations with the following parameter values: \emph{r} = 1; \emph{K}
= 4; \emph{e} = 0.8; \emph{m} = 1; \emph{A} = 2; \emph{h} = 0.4; and
\(w_{11} = w_{22} = 0.6\). I set an initial value of
\(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and \(a_{21}\) depended on
the value of \emph{n}. I set initial consumer and resource abundances
to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All simulations were conducted
in R \citep{R} and the code to reproduce these simulations is publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}).

\begin{figure}
\centering
\includegraphics{attack_tradeoffs.pdf}
\caption{\label{fig:tradeoff}\textbf{Evolutionary tradeoffs in consumer
attack rates.} In each foraging scenario, I explored the effects of
three different tradeoffs: intermediate combinations of attack rates
(\emph{a}\textsubscript{1,\emph{j}}, \emph{a}\textsubscript{2,\emph{j}})
are higher than the extremes (green line, \emph{n} \textgreater{} 1);
extreme combinations of attack rates are higher than intermediate
investments (orange line, \emph{n} \textless{} 1); and all combinations
of attack rates have the same total attack rate (blue line, \emph{n} =
1). Points corresponding to attack rates at the beginning of the
simulation for \emph{C}\textsubscript{1}, which was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{12}).}
\end{figure}

\section{Results}\label{results}

\subsection{Resources occur in same
habitat}\label{resources-occur-in-same-habitat}
